Challenges in Developing Great MCQMC Software

### 2022-11-01 MCQMC2022ChallengesQMCSoftwareArticle.tex
Office Computer Updates

% MCQMC 2022 article on the Challenges of Great MCQMC Software
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox,footinfo]{svmult}

\smartqed
\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are
% not available on your system
\usepackage{graphicx}       % standard LaTeX graphics tool
% when including figure files

\usepackage{array,colortbl}
\usepackage{amsmath,amsfonts,amssymb,bm} % no amsthm, Springer defines Theorem, Lemma, etc themselves
%\usepackage[mathx]{mathabx}
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
	<5> <6> <7> <8> <9> <10>
	<10.95> <12> <14.4> <17.28> <20.74> <24.88>
	mathx10
}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareFontSubstitution{U}{mathx}{m}{n}
\DeclareMathAccent{\widecheck}{0}{mathx}{"71}

% Note that Springer defines the following already:
%
% \D upright d for differential d
% \I upright i for imaginary unit
% \E upright e for exponential function
% \tens depicts tensors as sans serif upright
% \vec depicts vectors as boldface characters instead of the arrow accent
%
% Additionally we throw in the following common used macro's:
\input{macros}

% Macros below are now included in macros.tex from MCQMC 2016 web site
% This spot formerly included macros that are now in macros.tex

% indicator boldface 1:
\DeclareSymbolFont{bbold}{U}{bbold}{m}{n}
\DeclareSymbolFontAlphabet{\mathbbold}{bbold}
%\newcommand{\ind}{\mathbbold{1}}

\usepackage{microtype} % good font tricks

\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\urlstyle{same}
\usepackage{bookmark}
\pdfstringdefDisableCommands{\def\and{, }}
\makeatletter % to avoid hyperref warnings:
\providecommand*{\toclevel@author}{999}
\providecommand*{\toclevel@title}{0}
\makeatother

\usepackage{bbm,mathtools,array,longtable,booktabs,graphicx,color,enumitem,xspace}
%\input FJHDef.tex

\newcommand{\QMCPYnorm}[2][{}]{\ensuremath{\left \lVert #2 \right \rVert}_{#1}}
\newcommand{\QMCPYnormnorm}[2][{}]{\ensuremath{\lVert #2 \rVert}_{#1}}
\newcommand{\QMCPYbignorm}[2][{}]{\ensuremath{\bigl \lVert #2 \bigr \rVert}_{#1}}
\newcommand{\QMCPYBignorm}[2][{}]{\ensuremath{\Bigl \lVert #2 \Bigr \rVert}_{#1}}
\newcommand{\QMCPYabs}[1]{\ensuremath{{\left \lvert #1 \right \rvert}}}
\newcommand{\QMCPYbigabs}[1]{\ensuremath{{\bigl \lvert #1 \bigr \rvert}}}

\providecommand{\HickernellFJ}{Hickernell}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\newcommand{\frednote}[1]{  {\textcolor{red}  {\mbox{**Fred:} #1}}}
\newcommand{\yuhannote}[1]{ {\textcolor{violet}  {\mbox{**Yuhan:} #1}}}
\newcommand{\tonynote}[1]{ {\textcolor{orange}  {\mbox{**Tony:} #1}}}

%\journal{Journal of Complexity}

\allowdisplaybreaks[4]

%   python styling
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{xcolor}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\lstdefinestyle{Python}{
    showstringspaces=false,
    language        = Python,
    basicstyle      = \small\ttfamily,
    morekeywords = {as},
    keywordstyle    = \color{blue},
    stringstyle     = \color{darkgreen},
    commentstyle    = \color{darkgreen}\ttfamily,
	breaklines = true,
	postbreak=\text{$\hookrightarrow$\space},
	% style >>> and ... 
	%   see: https://tex.stackexchange.com/questions/326655/make-a-keyword-in-listings-enviorment
	alsoletter = {>,.} ,
    morekeywords = [2]{>>>,...},
    keywordstyle = [2]\color{cyan}\bfseries}

\newcommand{\AGSComment}[1]{{\color{cyan} Aleksei: #1}}
\newcommand{\FJHComment}[1]{{\color{magenta} Fred: #1}}

\newcommand{\hmu}{\widehat{\mu}}
\newcommand{\IID}{\textup{IID}}
\newcommand{\LD}{\textup{LD}}
\newcommand{\IMP}{\textup{imp}}
\newcommand{\unif}{\textup{unif}}
\newcommand{\IIDsim}{\overset{\IID}{\sim}}
\newcommand{\LDsim}{\overset{\LD}{\sim}}
\newcommand{\dig}{\textup{dig}}
\newcommand{\mA}{\mathsf{A}}
\newcommand{\mSigma}{\mathsf{\Sigma}}
\newcommand{\mLambda}{\mathsf{\Lambda}}
\newcommand{\mI}{\mathsf{I}}

@@

\title*{Quasi-Monte Carlo Software}
\authorrunning{S.-C.\ T.\ Choi et al}
\author{Sou-Cheng Terrya Choi \and Fred J. Hickernell \and R. Jagadeeswaran \and Michael J. McCourt \and Aleksei G. Sorokin}
\institute{Sou-Cheng Terrya Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{schoi32@hawk.iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and
R. Jagadeeswaran \at
\and
Michael J. McCourt \at ???
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{This is the article based on my MCQMC 2020 Tutorial. \AGSComment{Latest code demo at: \href{https://colab.research.google.com/drive/1wijwt0ugsJ9chUHtzSb3lQpc3-N8mCDs?usp=sharing}{https://colab.research.google.com/drive/1wijwt0ugsJ9chUHtzSb3lQpc3-N8mCDs?usp=sharing}}.}

\end{document}
\section{Introduction} \label{sec:intro}
Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases QMC  achieves one hundredth of the error of IID MC in the same amount of time. Often, these efficiency gains are obtained simply by  replacing IID sampling by low discrepancy (LD) sampling, which is the heart of QMC.

>>>>

% MCQMC 2022 article on the Challenges of Great MCQMC Software
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox]{svmult}

% choose options for [] as required from the list
% in the Reference Guide

\usepackage{type1cm}        % activate if the above 3 fonts are
% not available on your system
%
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool
% when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom

\usepackage{newtxtext}       % 
\usepackage[varvw]{newtxmath}       % selects Times Roman as basic font

% see the list of further useful packages
% in the Reference Guide

%\makeindex             % used for the subject index
% please use the style svind.ist with
% your makeindex program

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%Fred start
\usepackage{url, xspace}
\providecommand{\HickernellFJ}{Hickernell\xspace}

%%%%Fred end

@@

\title*{Challenges in Developing Great MCQMC Software}
\authorrunning{S.-C.\ T.\ Choi et al}
\author{Sou-Cheng Terrya Choi \and Yuhan Ding \and Fred J. Hickernell \and Aleksei G. Sorokin}
\institute{Sou-Cheng Terrya Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{schoi32@hawk.iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and Yuhan Ding \at
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{}

\section{Introduction}

	\begin{itemize}
	\item Correct, e.g., not omit the zeroth point in a low discrepancy sequence \cite{Owe22a, scipySobol2020a}

	\item  Complete---contain the components or  easily access components in other libraries to solve real, complex problems

	\item  Accessible---downloadable; tutorials, demos, discussion forums, etc.\ for (new) users; written in a language that potential users speak; provide a consistent user interface

	\item Efficient---in terms of compute time and memory

	\item Current---include the latest and best algorithms

	\item  Sustainable---have a sufficient user base and developer community for updates and maintenance

	\item Scalable---take advantage of advanced computer architectures for speed and to solve large problems

\end{itemize}

	\begin{itemize}
	\item \cite{LEc2017a} provides a history of random number generation
	\item Notorious RANDU from the 1960s \cite{RANDU}, which fails the spectral test
	\item VEGAS \cite{Lep78a, Lep21a} importance sampling Monte Carlo algorithm favored by physicists
	\item ACM low discrepancy point generators \cite{BraFox88,BraFoxNie92,HonHic00a}
	\item FinDer \cite{PasTra95,FinDer} and BRODA \cite{BRODA20a} targeting quantitative finance
	\item Korobov cubature and scrambled Sobol' generators in NAG \cite{NAG27} for decades		
	\item Scrambled Sobol' and Halton generators in MATLAB \cite{MAT9.13} since 2008, and fixed a few years later
	\item BUGS \cite{BUGSBook, BUGSweb} and Stan \cite{STAN} for Markov Chain Monte Carlo (MCMC)
	\item LatMRG \cite{LEcCou97}, RNGStreams \cite{LEcEtal02},  SSJ \cite{LEc2002a,SSJ}, and LatNetBuilder \cite{LatNet} 
	\item Low discrepancy generators \cite{FriKel02,FriKelweb}, SamplePack \cite{SamplePack}, \cite{GruWeb},  and MatBuilder \cite{paulin2022} 
	\item Sobol' direction numbers \cite{JoeKuo03,JoeKuo08,SobolDirection}
	\item Fast CBC, Magic Point Shop, QMC4PDE and other code since 2004  at \cite{NuyWeb} 
	\item Multi-level software \cite{GilesSoft,GilesQSoft} 
	\item Data-driven error bounds and stopping criteria in GAIL \cite{ChoEtal21a} and QMCPy \cite{QMCPy2020a, ChoEtal22a}  
	\item Uncertainty quantification libraries Dakota \cite{DakotaUsersManual}, UQTk \cite{DebEtal04,UQTk}, and MUQ \cite{MUQ}  have some basic level low discrepancy sampling
	\item QMC framework in Julia since 2019 \cite{QMCJulia} by Robbe and others 
	\item Scrambled Sobol' in SciPy \cite{virtanen2020scipy} and PyTorch \cite{paszke2019pytorch} since several years ago
	\item TensorFlow QMC framework \cite{tfqfQMC2021a} since a year or so ago
\end{itemize}

We have stand-alone algorithms, small libraries, and pieces of larger libraries

\bibliographystyle{spmpsci.bst}
\bibliography{FJH23,FJHown23,QMCSoftwareArticle,sc}

\begin{frame}[allowframebreaks]{Selective History of Publicly Available MC\textbf{QMC} Software}
\vspace{-2ex}
	\begin{itemize}
		\item \cite{LEc2017a} provides a history of random number generation
		\item Notorious RANDU from the 1960s \cite{RANDU}, which fails the spectral test
		\item VEGAS \cite{Lep78a, Lep21a} importance sampling Monte Carlo algorithm favored by physicists
		\item ACM low discrepancy point generators \cite{BraFox88,BraFoxNie92,HonHic00a}
		\item FinDer \cite{PasTra95,FinDer} and BRODA \cite{BRODA20a} targeting quantitative finance
		\item Korobov cubature and scrambled Sobol' generators in NAG \cite{NAG27} for decades		
		\item Scrambled Sobol' and Halton generators in MATLAB \cite{MAT9.13} since 2008, and fixed a few years later
		\item BUGS \cite{BUGSBook, BUGSweb} and Stan \cite{STAN} for Markov Chain Monte Carlo (MCMC)
		\item LatMRG \cite{LEcCou97}, RNGStreams \cite{LEcEtal02},  SSJ \cite{LEc2002a,SSJ}, and LatNetBuilder \cite{LatNet} 
		\item Low discrepancy generators \cite{FriKel02,FriKelweb}, SamplePack \cite{SamplePack}, \cite{GruWeb},  and MatBuilder \cite{paulin2022} 
		\item Sobol' direction numbers \cite{JoeKuo03,JoeKuo08,SobolDirection}
		\item Fast CBC, Magic Point Shop, QMC4PDE and other code since 2004  at \cite{NuyWeb} 
		\item Multi-level software \cite{GilesSoft,GilesQSoft} 
		\item Data-driven error bounds and stopping criteria in GAIL \cite{ChoEtal21a} and QMCPy \cite{QMCPy2020a, ChoEtal22a}  
		\item Uncertainty quantification libraries Dakota \cite{DakotaUsersManual}, UQTk \cite{DebEtal04,UQTk}, and MUQ \cite{MUQ}  have some basic level low discrepancy sampling
	    \item QMC framework in Julia since 2019 \cite{QMCJulia} by Robbe and others 
		\item Scrambled Sobol' in SciPy \cite{virtanen2020scipy} and PyTorch \cite{paszke2019pytorch} since several years ago
		\item TensorFlow QMC framework \cite{tfqfQMC2021a} since a year or so ago
	\end{itemize}

We have stand-alone algorithms, small libraries, and pieces of larger libraries
\end{frame}

\begin{frame}{Why Do We Need Software?}
	\vspace{-4ex}
	\begin{itemize}
		\item Theory 
		\begin{itemize}
			\item Explains, justifies, and leads to algorithms
			\item Assumptions$\implies$success can be viewed as failure$\implies$what went wrong
		\end{itemize}
		\item Software 
		\begin{itemize}
			\item Makes algorithms practical
			\item Solves (new) applications
			\item Eliminates do-it-yourself
		\end{itemize}
		\item Applications 
  		\begin{itemize}
			\item Make societal impact
			\item Inspire new theory
		\end{itemize}

	\end{itemize}

\uncover<2->{The US Department of Energy is investing in studying how to develop great scientific software, including a recent \emph{Workshop on the Science of Scientific-Software Development and Use} \cite{Her19a,ASCR-SSSDU}}
\end{frame}

\section{Structure}

\begin{frame}{Software Architecture}
	\vspace{-6ex}

\[
\mu :=  \int_{\ct} g(\vt) \, \lambda(\vt) \, \dif \vt = \cdots = 	
\underbrace{\Ex[f(\vX)]}_{\text{expectation}} = \underbrace{\int_{[0,1]^d}  f(\vx) \, \dif \vx}_{\text{integration}} \approx  \frac 1{n} \sum_{i=1}^{n} f(\vX_i) =: \hmu_{n}
\]

\vspace{-3ex}

\begin{description}[alabel]
	%\setlength{\labelwidth}{5ex}

	\item[Low Discrepancy Generator] producing $\{\vX_1, \vX_2, \dots \}$ that mimics the uniform distribution

	\item[True Measure] that defines the original integral, e.g., Lebesgue; embodies the transformation $\vt = \vPsi(\vx)$

	\item[Integrand] $g$, which defines the original integral, plus the transformed version, $f$, to fit the LD generator

	\item[Stopping Criterion] based on a data-driven error bound, which determines how large $n$ should be to ensure that $\abs{\mu - \hmu_n} \le \varepsilon$
\end{description}

\vspace{-1ex}
\uncover<2->{May we agree on a common software definition for these these objects like we have for floating point numbers or basic mathematical functions?}
\end{frame}

\begin{frame}{Choosing a Language, Library, or Environment}
	\vspace{-5ex}
	\begin{itemize}
    \item Language/library choices of the users are driven by familiarity and speed

  	\item  Favorites aligned with various communities, but no one dominates all

	\item Difference between prototyping environments and ninja written production code

    \item Growth of multi-processor environments 
    \begin{itemize}
    	\item Provides opportunity
    	\item Complicates software development; where does parallelism really help?
    \end{itemize}

    \item<2-> Recommendations if you want to contribute software
    \begin{itemize}
    \item Realize your new idea as a demo in your favorite library
    \item Add a significant idea as a feature to other libraries
    \item Move basic features of your library to larger library
    \item Write wrappers or demos to connect other libraries to your library

    \end{itemize}
 \end{itemize}
\end{frame}

\begin{frame}{Ex.\ QMCPy $+$ FEniCS/Dolfin \cite{LoggEtal_10_2012}}
\vspace{-3ex}
	\begin{itemize}
		\item Want to demonstrate how our QMC software could work with a popular DE solver for DEs with random coefficients
		\item Partial success \only<3->{Why?}
	\end{itemize}
 	\only<1>{\vspace{-3ex}}\only<2>{\vspace{-2ex}}\only<3>{\vspace{0ex}}
 \begin{minipage}{0.48\textwidth}
 	\only<1-2>{\begin{gather*}
 		-\frac{\dif }{\dif x}\biggl(a(x,\vW) \frac{\dif}{\dif x} u(x,\vW) \biggr) = 200( 2x - 1)\\
 		 0 \leq x \leq 1, \qquad
 		u(0) = u(1) = 0 \\
 		a(x,\vW) = 1 + 0.6 \sum_{k=1}^d \frac{W_k T_{k}(2x-1)}{k^2} \\
 		 \vW \sim \cu[-1,1]^d \\
 		 \Ex[u(0.25,\vW)] = \only<1>{\, ?}\only<2>{ -1.7462}
 	\end{gather*}}
  \only<3->{
\begin{itemize}
\item Even with FEniCS/Dolfin's extensive documentation we had difficulties
\begin{itemize}
    \item Did not understand the latest version, so reverted to an older version
    \item Had to learn how to change the random instance of the coefficient without tiggering the JIT compiler
    \item Do not yet know how to express the random coefficient in terms of the covariance kernel
\end{itemize}
\end{itemize}
  }
 \end{minipage}%
\hfill
 \begin{minipage}{0.48\textwidth}
 	\only<1-2>{\centering
 	\includegraphics<1>[width=\textwidth]{ProgramsImages/axwtrue0.6.eps}
 	\includegraphics<2->[width=\textwidth]{ProgramsImages/timing-4.eps}}%
  \only<3->{
\begin{itemize}
\item But
\begin{itemize}
    \item It  works
    \item If we can overcome the challenges, will post a blog and demo at \cite{QMCBlog}
\end{itemize}
\end{itemize}

  }
\end{minipage}

\end{frame}

\section{Support}
\begin{frame}{Encouraging Users of Your Library Requires}

	\vspace{-4ex}
	\begin{itemize}
		\item A repository where
		\begin{itemize}
		\item  Your library can be easily downloaded
		\item There is documentation
		\item Issues or queries can be posted
		\item There are updates
		\item Tests are run regularly to mitigate against bugs
		\end{itemize}
		\item (Elementary) demos on how to use the software and highlighting  its advantages
		\item Demos connecting  your library with others
		\item Recommending your students and postdocs to use your library
		\item Welcoming instructions on how to contribute an algorithm or a demo
	\end{itemize}

\end{frame}

\begin{frame}{Building a Developer Team for Your Library}

	\vspace{-4ex}
\begin{itemize}

\item Many senior researchers don't code, and rely on transient team members (students, postdocs)
\begin{itemize}
	\item Encourage a sense of ownership that lasts beyond the time your team members are part of your lab
	\item Encourage those outside your own group to contribute
\end{itemize}

\item Those of us who understand theory and appreciate software need to encourage and steer popular software to ensure that it is implemented well.

\item Recognize good software as valuable scholarly output

\item Recognize research software engineer as a valued vocation; see Research Software Engineers International 	\url{https://researchsoftware.org}, \emph{Research Software Engineers are people who combine professional software expertise with an understanding of research. They go by various job titles but the term Research Software Engineer (RSE) is fast gaining international recognition.}
  \end{itemize}
\end{frame}

\section{Next Steps}

\begin{frame}{Next Steps}

	\vspace{-2ex}
	\begin{minipage}{0.48\textwidth}
		\begin{itemize}
			\item Which exemplary theoretical developments are not yet implemented in readily available software?
			\item What potential MCQMC user communities are ill-served by existing  software libraries?
			\item Should those of us interested in software gather regularly for show-and-tell to foster more collaboration?
		\end{itemize}
	\end{minipage}
	\begin{minipage}{0.48\textwidth}
		Join us for a lunch discussion at Bella Casa tomorrow, July 19, at 12:45 PM

		\vspace{2ex}

		\includegraphics[width=\textwidth]{ProgramsImages/PizeriaBellaCasa.png}
\end{minipage}

\end{frame}

\begin{frame}[allowframebreaks]{References}
	\printbibliography
\end{frame}

\section{Introduction} \label{sec:intro}
Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases QMC  achieves one hundredth of the error of IID MC in the same amount of time. Often, these efficiency gains are obtained simply by  replacing IID sampling by low discrepancy (LD) sampling, which is the heart of QMC.
